Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-05-19 document/ms.tex
a first pass at the introduction

\newcommand{\modela}{\emph{Model A}}
\newcommand{\modelb}{\emph{Model B}}

@@

CITATIONS:
\begin{itemize}
\item
\citet{fressin-fp} --- simultaneous occurence rate and false positive rate
\item
\citet{morton} --- false positive rate
\item
\citet{morton-swift} --- KDE estimator for radius distribution of cool KOIs
\item
\citet{howard} --- inv-det-eff for short-period candidates around Sun-like
stars
\item
\citet{dressing} --- inv-det-eff for cool KOIs
\item
\citet{dong} --- semi-analytic completeness, likelihood function-ish,
turn-over at small radii, non-separability
\item
\citet{tabachnik} ---- Poisson likelihood
\item
\citet{tremaine} --- statistics of multis
\item
\citet{youdin} --- Poisson likelihood
\item
\citet{catanzarite} --- extrapolation to Eta-Earth assuming complete
\item
\citet{traub} --- extrapolation to Eta-Earth assuming complete
\end{itemize}

\hoggtodo{%
\kepler\ has been busting out exoplanets (CITE).
Many of them even appear possibly rocky and habitable (CITE).
Nonetheless, none yet known are true Earth analogs.
}

\hoggtodo{%
This hasn't stopped anyone from estimating the rate at which Sun-like stars
host Earth-like planets on year-ish period orbits (CITE).
All of these rate estimates are extrapolations of one kind or another.
Here we perform an extremely conservative extrapolation of this kind.
}

\hoggtodo{%
What would make an extrapolation like this conservative?
Using \emph{huge bins} does not a conservative estimate make.
You don't want to assume separabilities you can't justify.
You want to explore an enormous function space for the exoplanet rates or pdf.
We are going to do be conservative in these senses here, far more conservative
than anyone who has come before.
}

\dfmtodo{%
Cite \citet{tremaine} for likelihood function including multiplicity.
}

@@

We generate synthetic catalogs from two qualitatively different rate
functions.
Both distributions are generated by a separable model
\begin{eqnarray}
\rate_\ratepars (\ln\period,\,\ln\radius) &=&
    \rate_\ratepars^{(\period)}(\ln\period)\,
    \rate_\ratepars^{(\radius)}(\ln\radius)
\end{eqnarray}
but fit using the full general model.
The first model---\modela---has a smooth occurence surface where both
distributions are smooth broken power laws.
The second---\modelb---is designed to be exactly the distribution inferred by
\citet{petigura} in the range that they considered and then smoothly
extrapolated outside that range.
The catalogs generated from these two models are shown in \fig{smooth-results}
and \fig{simulation-results}, respectively and the data are available online
at \dfmtodo{ADD SOME URL}.

>>>>

\newcommand{\modela}{\emph{Catalog A}}
\newcommand{\modelb}{\emph{Catalog B}}

@@

% CITATIONS:
% \begin{itemize}
% \item
% \citet{fressin-fp} --- simultaneous occurence rate and false positive rate
% \item
% \citet{morton} --- false positive rate
% \item
% \citet{morton-swift} --- KDE estimator for radius distribution of cool KOIs
% \item
% \citet{howard} --- inv-det-eff for short-period candidates around Sun-like
% stars
% \item
% \citet{dressing} --- inv-det-eff for cool KOIs
% \item
% \citet{dong} --- semi-analytic completeness, likelihood function-ish,
% turn-over at small radii, non-separability
% \item
% \citet{tabachnik} ---- Poisson likelihood
% \item
% \citet{tremaine} --- statistics of multis
% \item
% \citet{youdin} --- Poisson likelihood
% \item
% \citet{catanzarite} --- extrapolation to Eta-Earth assuming complete
% \item
% \citet{traub} --- extrapolation to Eta-Earth assuming complete
% \end{itemize}

NASA's \kepler\ mission has enabled the discovery of thousands of exoplanet
candidates (\dfmtodo{CITE KEPLER}).
Many of these planets orbit Sun-like stars (\citealt{petigura}), where the
definition of Sun-like is given in terms of the star's temperature and surface
gravity.
Given these catalogs, it is interesting to ask what we can say about the
occurence rate of exoplanets as a function of their physical parameters
(period, radius, \etc).
The two main reasons why this is interesting are that \emph{(a)} theories of
planet formation make predictions about the shape of this function
(\dfmtodo{CITE}), and \emph{(b)} it allows us to place probabilistic
constraints on the rate of Earth analogs\footnote{We define, in detail, what
we mean by Earth analog below but, in short, we mean Earth-sized planets on
year-long orbits around Sun-like stars.}.

\citet{petigura} recently published constraints on this function based on
their independent analysis of the \kepler\ light curves for 42,000 Sun-like
stars but the study of exoplanet occurence rates has a long history.
\dfmtodo{Summarize history with lotsa citations.}

A very important component of any study of exoplanet occurence rates is the
treatment of detection efficiency.
Qualitatively, in a transit survey, small planets with long periods are much
harder to detect than large planets orbiting close to their star.
This effect is degenerate with any inferences about the rate function and it
can be hard to constraint quantitatively.
In practice, there are three methods for taking this effect into account:
\emph{(a)} making conservative cuts on the candidates and assuming that the
resulting catalog is complete (\citealt{catanzarite, traub}; \dfmtodo{CITE}),
\emph{(b)} asserting an analytic form for the detection efficiency as a
function of approximate signal-to-noise (\citealt{howard, dressing, dong,
fressin-fp, morton-swift}; \dfmtodo{CITE}), and \emph{(c)} determining the
detection efficiency empirically by injecting synthetic signals into the raw
data and testing recovery rates (\citealt{petigura-a, petigura}).

In this \paper, we develop a flexible framework for probabilistic inference of
exoplanet occurence rates that can be applied to incomplete catalogs with
\emph{non-negligible observational uncertainties}.
Our method is a generalization of a previously studied method
(\citealt{hogge}) to account for detection efficiencies.
We run tests on simulated datasets---comparing results with the standard
techniques that neglect observational uncertainties---and apply our method to
a real catalog of small planets transiting Sun-like stars
(\citealt{petigura}).
For the purposes of this \paper, we make some strong assumptions---although we
argue that they are weaker than the implicit assumptions in previous
studies---but none of these are necessary for the applicability of our method.
Specifically, we make the following assumptions
\begin{itemize}

\item the candidates in the catalog are independent Poisson draws from the
censored occurence rate,

\item every candidate is a real exoplanet (there are no false positives),

\item the observational uncertainties on the physical parameters are
non-negligible but the amplitudes are known (the catalog provides
probabilistic constraints on the parameters),

\item the detection efficiency of the pipeline is known, and

\item the true occurence rate function is \emph{smooth}\footnote{We give our
definition of ``smooth'' in more detail below but our model is very flexible
so this is not a strong restriction.}.

\end{itemize}
The first assumption---conditional independence of the candidates---is
reasonable since the dataset that we consider explicitly includes only single
transiting systems (\citealt{petigura}).
When considering other datasets it will be interesting and important to study
the distributions of taking multiple transiting systems into account
(\dfmtodo{CITE}) and this would be possible using a conditional distributions
(as developed by \citealt{tremaine}).
The second assumption---neglecting false positives---is also strong and not
really justified (\citealt{fressin-fp, morton}).
For this \paper, we will neglect this issue and only comment on the effects
but the prior distributions published by \citet{fressin-fp} would be directly
applicable in our method and this would be a very interesting generalization.

@@

We generate synthetic catalogs from two qualitatively different rate
functions.
Both distributions are generated by a separable model
\begin{eqnarray}
\rate_\ratepars (\ln\period,\,\ln\radius) &=&
    \rate_\ratepars^{(\period)}(\ln\period)\,
    \rate_\ratepars^{(\radius)}(\ln\radius)
\end{eqnarray}
but fit using the full general model.
The first catalog---\modela--- is generated assuming a smooth occurence
surface where both distributions are broken power laws.
The second---\modelb---is designed to be exactly the distribution inferred by
\citet{petigura} in the range that they considered and then smoothly
extrapolated outside that range.
The catalogs generated from these two models are shown in \fig{smooth-results}
and \fig{simulation-results}, respectively and the data are available online
at \dfmtodo{ADD SOME URL}.
